### 2018-07-18 source/Ch2-techniques/imaging.tex
final commit for sake of backup - thesis completed

\pg
Astronomers who request observation time on interferometric arrays are generally not experts in the theory and operation of these arrays. This is a simple consequence of scientific division of labour. This thesis is written from the perspective of an ``expert" user: one who specialises in understanding interferometric arrays and reducing their data, rather than their subsequent analysis. However, it can be very helpful to contextualise this perspective in the experience of other scientists, which is what this section aims to do. This is simply to highlight the conceptual structure of a radio interferometric observation, and where the concepts outlined in this chapter fall in the overall pattern. In other words, it is best understood as a guide to see where a given section's topic slots into the overall production of an image made from visibilities.
\pg
LOFAR data can be acquired through observation or via the Long-Term Archive. The worst of this data will already be flagged. Users can thus generally proceed directly to calibration. Because there is no guarantee of sufficient signal-to-noise in the ``target field" (the object that the astronomer is actually interested in), it is standard practice to observe a calibrator source for 5 minutes before and after each observation. Such sources typically need to be bright and unresolved for the interferometer being used. Because they are bright and at phase centre for the 5-minute observation, the SNR when solving for calibration solutions for these 5 minutes will tend to be quite good. These solutions will then be interpolated between the 5 minutes onto the 8-hour observation. Calibration will be described in \cref{section.calibration}.

\pg
So far, the astronomer will have worked entirely with visibilities. However, generally speaking, the aim will be to get information on the object's brightness distribution at the observing frequency. This will require \emph{imaging}: going from visibilities to the image-plane. This can be done through a variety of tools, but will generally involve deconvolution. Imaging will be the subject of the rest of this chapter, and deconvolution will be described in more detail in \cref{section.clean}.

@@

\pg
There are two quantities of interest to all astronomers: sensitivity and angular resolution\footnote{Other kinds of resolutions - in time and frequency, for example - are also extremely important to many astronomers, but are not what concern us here.}. An instrument's sensitivity is a function of its collecting area\footnote{It is also a function of technological factors, of course, but \emph{ceteris paribus}, a more sensitive telescope means a telescope with a wider collecting area}. Resolution, for well-designed instruments\footnote{By this, we mean that we assume that an instrument is also designed to optimise resolution.}, is limited by diffraction in the absence of atmospheric effects. This introduces specific issues in the radio domain. Radio waves have very long wavelengths - often comparable to meters, rather than the $\sim100-1000$nm wavelengths of optical light. Achieving a resolution comparable to those of optical telescopes would thus require making telescopes with apertures tens of millions of times larger than those already titanic instruments!

@@

\pg
We have defined what a baseline corresponds to: a vector coordinate in $(u,v,w)$-space. To each baseline we associate a measurement, which we call the \emph{visibility}. 
\begin{figure}[ht]
\centering
\includegraphics[width=.7\textwidth]{images/visibility-creation.png}
\caption{\label{fig.visibility} There are two ways to combine the voltages from two antennas into a visibility: they are sum-correlation and $\pi$-correlation. In this manuscript, we will only concern ourselves with the latter. Image credit: Julien Girard}
\end{figure}
The visibility associated with baseline $\mathbf{b}_{AB}$ is created by taking the voltage $V_A$ measured by antenna A, multiply it by the complex conjugate $V_B^*$ of the voltage measured by antenna B, average over the correlator dump time (i.e. the time over which the measurement is made). This scalar quantity is associated to the baseline position vector $\mathbf{b}_{AB}$. In other words:
\begin{align}
\mathbf{b}_{AB} &= \frac{\mathbf{x}_{B}-\mathbf{x}_{A}}{\lambda_\mathrm{obs}}\\
V_{AB}          &= \langle V_{A} V_{B}^*\rangle_{\delta t, \delta \nu} 
\end{align}
where $\langle \cdots \rangle_{i}$ denotes an averaging over quantity $i$. We see that a visibility is a complex vector quantity. We also see that $\mathbf{b}_{AB} = \mathbf{b}_{BA}^*$: the information of the visibility associated with baseline BA is contained in the visibility associated with baseline AB. This means that in practice, only half of the visibilities ever need be stored. What does the visibility measurement correspond to?
\begin{figure}[ht]
\centering
\includegraphics[width=.5\textwidth]{images/visibility-measure.png}
\caption{\label{fig.visibility.measure} Here, we assume that there are only two sources in the sky whose signal can be measured by our antennas. The final visibility is the sum of the visibilities associated with each individual source. Image credit: Julien Girard \textcolor{red}{add reference to julien's phd}}
\end{figure}

\pg
Different astrophysical sources are located very far apart, and so their signals do not interfere. They are measured additively in both antennas. Provided that the signal from both sources is coherent when observed by the dishes, the correlation between the voltages measured by antennas A and B will simply be the sum of the voltage correlations associated with individual sources. This is shown in \cref{fig.visibility}: the radio signal from two different sources, at two different locations, is picked up simultaneously by both radio antennas. Because only the correlation between both antennas is measured, the signals do not interfere: if the signal from each source gives a voltage $V_{p}^s$ for antenna $p$, then the additive property of electric fields means that the total visibility measured by baseline $pq$ can be written as:
\begin{align}
V_{pq}  &= \langle \left(\sum_{s} V_{p}^s\right) \left(\sum_{s} (V_{q}^s)\right)^*  \rangle_{\Delta t} \\
        &= \langle \sum_{s,s'} V_{p}^s (V_q^{s'})^*  \rangle_{\Delta t}\\
        &= \langle \sum_s V_{p}^s (V_q^{s})^* \rangle_{\Delta t} + \underbrace{\langle \sum_{s,s'\ne s} V_{p}^s (V_q^{s'})^* \rangle_{\Delta t}}_{=0} \label{eq.spatial.incor}
\end{align}
where the fact that signal from different sources is not coherent means that we average out cross-contributions (i.e. signal from $s\ne s'$). 
%\begin{align}
%V_{pq}&=\langle \left(\sum_{s} V_{p}^s\right) \left(\sum_{s} V_{q}^s \right)^* \rangle_{\Delta t} \right)\\
%      &=\langle\sum_{s,s'} \left( V_p^s\left(V_q^{s'}\right)^* \rangle_{\Delta t}
%\end{align}

\pg
Note that in Fig. \ref{fig.visibility.measure}, neither source is at the zenith. This introduces the notion of the \emph{effective baseline} which is shorter than the \emph{physical baseline} when measuring a signal from a source that is not at the zenith. Sources in different positions in the sky will be measured by an interferometric array, but the same array will have different effective baselines for different sources. To point an interferometric array in a specific direction in the sky, one must introduce a phase delay in each antenna (or, for fundamental interferometer elements such as the LOFAR tiles or NenuFAR, by playing with the cable length between each dish and the correlator). Because the phase needs to be properly sampled to limit information loss, a process known as fringe stopping\footnote{see \href{https://casper.berkeley.edu/astrobaki/index.php/Fringe_Stopping}{here} for more information. Fringe stopping consists of measuring the beat between a predicted fringe rate (i.e. the period of the visibility sine wave corresponding to a point source out of phase centre; if the position of an array's antennas is well-known, this can be modeled accurately) and the actual fringe rate (i.e. the fringe rate corresponding to the true period of the visibility sine wave). The difference between the two is a function of how well the antennas' positions are known. If they are well-known, then the phase information can still be measured, even without very fine sampling in time. } must be used to avoid taking a measurement every few milliseconds. The position of sources are then given in terms of the cardinal angle $\sigma$: this is the difference between the position of a source and the phase centre, which is where the array is pointing. For example, in \cref{fig.visibility.measure}, if we point antennas and introduce phase delays such that we are pointing the array towards Source 2, the position of Source 1 will be $\sigma$.

@@

\pg
In general, interferometric design is such that the $w$ component of visibilities' $(u,v,w)$ coordinates is negligible (or can be put in a frame of reference where it can usually be approximated as such). Radio astronomers tend to thus talk of a $uv$-plane rather than $uvw$-space to describe visibility space. The set of $uv$-values for all the baselines of an interferometric array is known as its $uv$-coverage, and defines the array's properties entirely.
For the VLA, for example, the instantaneous $uv$-coverage when observing the zenith will be as shown in Fig. \ref{fig.vla.uvcoverage}.

@@

\pg
The problem of imaging radio interferometric data is still a very active field of research today. For more information on this topic, the NRAO website\footnote{See \href{https://www.cv.nrao.edu/~abridle/deconvol/node8.html\#SECTION00051000000000000000}{the CLEAN section of the NRAO website}.} is an excellent start for a general introduction to this topic, while \citet{2017arXiv171202078T} provides an outstanding entry point for cutting-edge work at the time of writing.

@@

\pg
The previous section mentions that good deconvolution is dependent on good calibration: poor calibration introduces artefacts in images, which can in turn bias deconvolution. A wrong model of the sky brightness distribution, in turn, will bias calibration, introducing artefacts in images. This feedback of one inverse problem onto the other is a strong indication that both are, in fact, different facets of the same underlying inverse problem: going from measured visibilities to a sky brightness distribution. Indeed, one could - conceptually, at least - solve for both problems at once: find both the sky brightness distribution and the antenna gains which best fit the measured visibilities. In practice, this is not feasible; not only is there no proof that the overall calibration \& imaging problem is convex, but even if it is, it is also poorly conditioned: multiple sets of sky brightness distributions \& antenna gains are equally viable for a given set of measured visibilities. Examples of convexity and conditioning are shown in \cref{fig.examples}.

@@

\pg
Because the net calibration \& imaging problem has so many degrees of freedom, it is convenient to break it into two parts: calibration and imaging. By doing this, instead of exploring a massive set of hyperparameters (e.g. relative flux distribution at a given frequency, overall spectral index, 8 values per antenna per time and frequency interval, ...), a smaller subset of these hyperparameters is explored while all else is kept constant. By ensuring that these subsets each give reasonable results individually, it becomes possible to solve the full problem. This practice of imposing specific priors to each subset of the problem is a form of regularisation: it can turn a poorly-conditioned problem into a better-conditioned one, provided that the priors are well-chosen.

@@

\pg
\cref{fig.briggs-weighting} clearly shows that imaging done with the same data but different weighting schemes give radically different deconvolved images. Images made with uniform (\cref{fig.uniform}) or Briggs (\cref{fig.briggs0}) weighing are much sharper, with uniform weighting (\cref{fig.uniform}) giving such resolution that the sources are nearly invisible in the field (though they are present - the contrast was not upped so as to show all three images on the same flux scale).

@@

>>>>

@@

\pg
There are two quantities of interest to all astronomers: sensitivity and angular resolution\footnote{Other kinds of resolution - in time and frequency, for example - are also extremely important to many astronomers, but are not what concern us here.}. An instrument's sensitivity is a function of its collecting area\footnote{It is also a function of technological factors, of course, but \emph{ceteris paribus}, a more sensitive telescope means a telescope with a wider collecting area}. Resolution, for well-designed instruments\footnote{By this, we mean that we assume that an instrument is also designed to optimise resolution.}, is limited by diffraction in the absence of atmospheric effects. This introduces specific issues in the radio domain. Radio waves have very long wavelengths - often comparable to meters, rather than the $\sim100-1000$nm wavelengths of optical light. Achieving a resolution comparable to those of optical telescopes would thus require making telescopes with apertures tens of millions of times larger than those already titanic instruments!

@@

\pg
We have defined what a baseline corresponds to: a vector coordinate in $(u,v,w)$-space. To each baseline we associate a measurement, which we call the \emph{visibility}. 
\begin{figure}[ht]
\centering
\includegraphics[width=.7\textwidth]{images/visibility-creation.png}
\caption{\label{fig.visibility} There are two ways to combine the voltages from two antennas into a visibility: they are sum-correlation and $\pi$-correlation. In this manuscript, we will only concern ourselves with the latter. Image credit: Julien Girard}
\end{figure}
The visibility associated with baseline $\mathbf{b}_{AB}$ is created by taking the voltage $V_A$ measured by antenna A, multiply it by the complex conjugate $V_B^*$ of the voltage measured by antenna B, average over the correlator dump time (i.e. the time over which the measurement is made). This scalar quantity is associated to the baseline position vector $\mathbf{b}_{AB}$. In other words:
\begin{align}
\mathbf{b}_{AB} &= \frac{\mathbf{x}_{B}-\mathbf{x}_{A}}{\lambda_\mathrm{obs}}\\
V_{AB}          &= \langle V_{A} V_{B}^*\rangle_{\delta t, \delta \nu} 
\end{align}
where $\langle \cdots \rangle_{i}$ denotes an averaging over quantity $i$. We see that a visibility is a complex vector quantity. We also see that $\mathbf{b}_{AB} = \mathbf{b}_{BA}^*$: the information of the visibility associated with baseline BA is contained in the visibility associated with baseline AB. This means that in practice, only half of the visibilities ever need be stored. What does the visibility measurement correspond to?
\begin{figure}[ht]
\centering
\includegraphics[width=.5\textwidth]{images/visibility-measure.png}
\caption{\label{fig.visibility.measure} Here, we assume that there are only two sources in the sky whose signal can be measured by our antennas. The final visibility is the sum of the visibilities associated with each individual source. Image credit: Julien Girard \citepads{julien}}
\end{figure}

\pg
The signals from different astrophysical do not interfere. They are measured additively in both antennas. Provided that the signal from each sources is coherent when observed by the dishes, the correlation between the voltages measured by antennas A and B will simply be the sum of the voltage correlations associated with individual sources. This is shown in \cref{fig.visibility}: the radio signal from two different sources, at two different locations, is picked up simultaneously by both radio antennas. Because only the correlation between both antennas is measured, the signals do not interfere: if the signal from each source gives a voltage $V_{p}^s$ for antenna $p$, then the additive property of electric fields means that the total visibility measured by baseline $pq$ can be written as:
\begin{align}
V_{pq}  &= \langle \left(\sum_{s} V_{p}^s\right) \left(\sum_{s} (V_{q}^s)\right)^*  \rangle_{\Delta t} \\
        &= \langle \sum_{s,s'} V_{p}^s (V_q^{s'})^*  \rangle_{\Delta t}\\
        &= \langle \sum_s V_{p}^s (V_q^{s})^* \rangle_{\Delta t} + \underbrace{\langle \sum_{s,s'\ne s} V_{p}^s (V_q^{s'})^* \rangle_{\Delta t}}_{=0} \label{eq.spatial.incor}
\end{align}
where the fact that signal from different sources is not coherent means that we average out cross-contributions (i.e. signal from $s\ne s'$). 
%\begin{align}
%V_{pq}&=\langle \left(\sum_{s} V_{p}^s\right) \left(\sum_{s} V_{q}^s \right)^* \rangle_{\Delta t} \right)\\
%      &=\langle\sum_{s,s'} \left( V_p^s\left(V_q^{s'}\right)^* \rangle_{\Delta t}
%\end{align}

\pg
Note that in Fig. \ref{fig.visibility.measure}, neither source is at the zenith. This introduces the notion of the \emph{effective baseline} which is shorter than the \emph{physical baseline} when measuring a signal from a source that is not at the zenith. Sources in different positions in the sky will be measured by an interferometric array, but the same array will have different effective baselines for different sources. To point an interferometric array in a specific direction in the sky, one must introduce a phase delay in each antenna (or, for fundamental interferometer elements such as the LOFAR tiles or NenuFAR, by playing with the cable length between each dish and the correlator). This is done by three means: delay tracking (which is exact only for the delay tracking centre\footnote{Source: \url{http://www.astron.nl/eris2013/Documents/2_laing_fundamentals_interferometry.pdf}, \url{https://science.nrao.edu/science/meetings/2016/15th-synthesis-imaging-workshop/SISS15Advanced.pdf}}, and corresponds to playing with the cable length between antennas in the interferometric array), antenna pointing\footnote{This maximises antenna sensitivity in a given direction, but is not done with LOFAR}, and fringe stopping\footnote{Fringe stopping consists of down-converting the signal received by antennas by mixing with a local oscillator. The geometric delay compensation is usually applied on this down-converted signal. Source: \url{http://www.gmrt.ncra.tifr.res.in/gmrt_hpage/Users/doc/WEBLF/LFRA/node74.html}}.
%Fringe stopping consists of measuring the beat between a predicted fringe rate (i.e. the period of the visibility sine wave corresponding to a point source out of phase centre; if the position of an array's antennas is well-known, this can be modeled accurately) and a known signal, e.g. the actual fringe rate (i.e. the fringe rate corresponding to the true period of the visibility sine wave). The difference between the two is a function of how well the antennas' positions are known. If they are well-known, then the phase information can still be measured, even without very fine sampling in time. } must be used to avoid taking a measurement every few milliseconds. 
This means that there are actually three different ``field centres", which are usually (but not always) set to be the same: the pointing (direction of maximum antenna sensitivity), the delay (which can be introduced in cable length) and the phase tracking (which consists of applying a known signal to the measured signal and recording the resulting heterodyne wave so as to keep a reasonable amount of post-correlation data without losing information on sources outside of the phase delay centre). The position of sources are then given in terms of the cardinal angle $\sigma$: this is the difference between the position of a source and the phase centre, which is where the array is pointing. For example, in \cref{fig.visibility.measure}, if we point antennas and introduce phase delays such that we are pointing the array towards Source 2, the position of Source 1 will be $\sigma$.

@@

\pg
%In general, interferometric design is such that the $w$ component of visibilities' $(u,v,w)$ coordinates is negligible (or can be put in a frame of reference where it can usually be approximated as such). 
For the sake of brevity, radio astronomers tend to talk of a $uv$-plane rather than $uvw$-space to describe visibility space. The set of $uvw$-values for all the baselines of an interferometric array is known as its $uv$-coverage, and defines the array's properties entirely.
For the VLA, for example, the instantaneous $uv$-coverage when observing the zenith will be as shown in Fig. \ref{fig.vla.uvcoverage}.

@@

\pg
The problem of imaging radio interferometric data is still a very active field of research today. For more information on this topic, the NRAO website\footnote{See \href{https://www.cv.nrao.edu/~abridle/deconvol/node8.html\#SECTION00051000000000000000}{the CLEAN section of the NRAO website}.} is an excellent start for a general introduction to this topic, while \citet{2017arXiv171202078T} provides an outstanding example of cutting-edge work at the time of writing.

@@

\pg
The previous section mentions that good deconvolution is dependent on good calibration: poor calibration introduces artefacts in images, which can in turn bias deconvolution. A wrong model of the sky brightness distribution, in turn, will bias calibration, introducing artefacts in images. This feedback of one inverse problem onto the other is a strong indication that both are, in fact, different facets of the same underlying inverse problem: going from measured visibilities to a sky brightness distribution. Indeed, one could - conceptually, at least - solve for both problems at once: find both the sky brightness distribution and the antenna gains which best fit the measured visibilities. In practice, this is not practical (though it is feasible); not only is there no proof that the overall calibration \& imaging problem is convex, but even if it is, it is also poorly conditioned: multiple sets of sky brightness distributions \& antenna gains are equally viable for a given set of measured visibilities. This means that, even if the problem is convex, it takes an extremely long time to converge properly. Examples of convexity and conditioning are shown in \cref{fig.examples}.

@@

\pg
Bad conditioning and non-convexity can be alleviated through regularisation: by specifying a well-chosen prior (e.g. for interferometry, using a good sky model for calibration), the underlying distribution of the solutions can be constrained to fall in a narrow, convex region. This is shown in \cref{fig.selfcal.regularise}
\begin{figure}[h!]
	\includegraphics[width=.6\linewidth]{images/regularise.png}
	\caption{\label{fig.selfcal.regularise} Plot showing the effect of regularisation on an ill-conditioned, non-convex problem. The $\chi^2$ fit is constrained to lie in a convex, well-conditioned region: it falls within the global minimum, at at one point only rather than somewhere in a valley.}
\end{figure}

\pg
Because the imaging problem (i.e. the problem of deconvolution) is linear, it can only be convex - though it is not necessarily well-conditioned. As for the calibration problem, experience strongly indicates that it is nearly convex \textit{and} ill-conditioned. For calibration convexity, Cyril Tasse (pers. comm.) has found that the Hessian matrix of the gain solutions is positive semidefinite, i.e. that all its eigenvalues are positive and non-zero except for one. This behaviour likely explains why self-calibration is, in the practical experience of radio-interferometric practicioners,  neither well-conditioned nor convex: if the problem is poorly regularised, then the same initial model may give different end results, and to truly verify the convergence property of self-calibration would require both a full Monte Carlo simulation and enough time to allow the very slowly-converging procedures to finish converging. It is not a practical thing to test rigorously.

\pg
Because the net calibration \& imaging problem has so many degrees of freedom, it is convenient to break it into two parts: calibration and imaging. By doing this, instead of exploring a massive set of hyperparameters (e.g. relative flux distribution at a given frequency, overall spectral index, the 8 values of a complex $2\times 2$ matrix per antenna per time and frequency interval, ...), a smaller subset of these hyperparameters is explored while all else is kept constant. By ensuring that these subsets each give reasonable results individually, it becomes possible to solve the full problem. This practice of imposing specific priors to each subset of the problem is a form of regularisation: it can turn a poorly-conditioned problem into a better-conditioned one, provided that the priors are well-chosen.

@@

\pg
\cref{fig.briggs-weighting} shows images done with the same data and the same imaging parameters bar weighting scheme choice. We can see that different weighting schemes give radically different deconvolved images. Images made with uniform (\cref{fig.uniform}) or Briggs (\cref{fig.briggs0}) weighing are much sharper, with uniform weighting (\cref{fig.uniform}) giving such resolution that the sources are nearly invisible in the field (though they are present - the contrast was not upped so as to show all three images on the same flux scale).

@@

\pg
Astronomers who request observation time on interferometric arrays are generally not experts in the theory and operation of these arrays. This is a simple consequence of scientific division of labour. This thesis is written from the perspective of an ``expert" user: one who specialises in understanding interferometric arrays and reducing their data, rather than their subsequent analysis. However, it can be very helpful to contextualise this perspective in the experience of other scientists, which is what this section aims to do. This is simply to highlight the conceptual structure of a radio interferometric observation, and where the concepts outlined in this chapter fall in the overall pattern. In other words, it is best understood as a guide to see where a given section's topic slots into the overall production of an image made from visibilities.

\pg
LOFAR data can be acquired through observation or via the Long-Term Archive. The worst of this data will already be flagged. Users can thus generally proceed directly to calibration. Because there is no guarantee of sufficient signal-to-noise in the ``target field" (the object that the astronomer is actually interested in), it is standard practice to observe a calibrator source for 5 minutes before and after each observation. Such sources typically need to be bright and unresolved for the interferometer being used. Because they are bright and at phase centre for the 5-minute observation, the SNR when solving for calibration solutions for these 5 minutes will tend to be quite good. These solutions will then be interpolated between the 5 minutes onto the 8-hour observation. Calibration will be described in \cref{section.calibration}.

\pg
So far, the astronomer will have worked entirely with visibilities. However, generally speaking, the aim will be to get information on the object's spatial brightness distribution at the observing frequency. This will require \emph{imaging}: going from visibilities to the image-plane. This can be done through a variety of tools, but will generally involve deconvolution. Imaging will be the subject of the rest of this chapter, and deconvolution will be described in more detail in \cref{section.clean}.

### 2018-07-18 source/Ch2-techniques/lofar.tex
final commit for sake of backup - thesis completed

\pg
LOFAR is a SKA pathfinder instrument, which means that it serves not only as a cutting-edge instrument in its own right, but does so with the explicit aim of serving as testing grounds for technologies \& techniques which could be usefully implemented in the SKA. It is in this context, for example, that trailblazers such as NenuFAR \citepads{2012sf2a.conf..687Z}, a low-frequency extension of LOFAR, are tested. LOFAR is an interferometric array, with each of its elements made up of antennas combined into a phase array. It thus consists of antennas which are combined to form stations, which are themselves distributed throughout the Netherlands and Europe.
\begin{figure}[h!] 
  \begin{minipage}[c]{0.45\linewidth}
    \centering
    \includegraphics[width=\linewidth]{images/superterp.png}
	\subcaption{\label{fig.lofar.superterp} LOFAR core, known as the Superterp.}
%    \vspace{4ex}
  \end{minipage}
  \hfill
  \begin{minipage}[c]{0.45\linewidth}
    \centering
    \includegraphics[width=.5\linewidth]{images/{lofar-core-map_grey.jpg}}
    \subcaption{\label{fig.lofar.core} Distribution of the so-called LOFAR "core" stations, which include the Superterp.}
%    \vspace{4ex}
  \end{minipage} 
  \begin{minipage}[c]{0.45\linewidth}
    \centering
    \includegraphics[width=\linewidth]{images/{Distribution_Remote_Stations}} 
    \subcaption{\label{fig.lofar.remote} LOFAR with both "core" and "remote" stations.}
%    \vspace{4ex}
  \end{minipage}
  \begin{minipage}[c]{0.45\linewidth}
    \centering
    \includegraphics[width=.9\linewidth]{images/{Distribution_International_Stations}} 
    \subcaption{\label{fig.lofar.international} International LOFAR. Newer stations (one in Ireland, three in Poland) are not shown here.}
%    \vspace{4ex}
  \end{minipage} 
\caption{\label{fig.lofar.distribution} Geographic location and distribution of LOFAR stations, explicitly showing what is meant by Superterp, core, remote and international stations. All images from \href{https://www.astron.nl/radio-observatory/astronomers/users/technical-information/lofar-array-configuration/lofar-array-conf}{the official ASTRON website}}
\end{figure}

@@

\pg
We see the presence, in both cases, of two very different antenna types. One of these antenna types is not a single antenna, but rather a phased array: 16 antenna dipoles distributed in a $4 \times 4$ array. International LOFAR stations include 48 such tiles, distributed in a ``filled-disk" pattern, where the disk fill ratio is determined by the size of the tile. This pattern is shown in \cref{fig.hba.tile}. Dutch ``remote" stations have 96 such tiles spread in 2 disks (HBA\textunderscore INNER and HBA\textunderscore OUTER), while ``core" station HBA tiles are split into two 24-tile crosses. The antennas from each tile are combined into a phased array with a single ``tile beam", and all tile beams are themselves combined as a phased array into a station beam\footnote{Referece: \href{https://www.astron.nl/radio-observatory/astronomers/technical-information/antennae/antennae-description}{ASTRON technical description}}. This station beam is then pointed digitally.
\begin{figure}[h!]
\includegraphics[width=0.45\textwidth]{images/{Schematic-diagram-of-a-24-tile-LOFAR-HBA-station-A-tile-is-made-of-16-dual-polarization.png}}
\caption{\label{fig.hba.tile} Layout of a single HBA tile. Source: \href{https://www.researchgate.net/profile/Sarod_Yatawatta/publication/281316394/figure/fig1/AS:614002589720576@1523401033149/Schematic-diagram-of-a-24-tile-LOFAR-HBA-station-A-tile-is-made-of-16-dual-polarization.png}{Sarod Yatawatta researchgate profile}}
\end{figure}

\pg
LOFAR-HBA is sensitive to higher frequencies, from 120 MHz to 240 MHz\footnote{Reference: \href{http://www.lofar.org/about-lofar/system/lofar-numbers/lofar-numbers}{LOFAR.org website}}. It has a smaller field of view than LOFAR-LBA, but a better angular resolution.

\pg
LBA antennas, meanwhile, follow a very simple - and cost-effective - design. A typical LBA antenna is shown in \cref{fig.lba.ant}. 96 such antennas are spread in a semi-random pattern in each LOFAR station\footnote{Reference: \href{http://www.lofar.org/about-lofar/system/lofar-numbers/lofar-numbers}{LOFAR.org website.}}. In Dutch stations, 48 of these are used at all times. They are combined as a single phased array.
\begin{figure}[h!]
\includegraphics[width=0.45\textwidth]{images/{lba.png}}
\caption{\label{fig.lba.ant} Picture of a single LBA antenna. Source: \href{https://i2.wp.com/lofar.ie/wp-content/uploads/2017/04/lba.png?resize=1500\%2C1000}{LOFAR technology website.}}
\end{figure}

\pg
The dipole design frees observers from the need to physically point antennas at all: the final station pointing is achieved by digitally introducing delays in observed phase before averaging the data. In this sense, the pointing is achieved in a similar way as individual HBA tile pointing, albeit in analog. These antennas are receptive to signals emitted in the 30-80 MHz frequency range.

>>>>

\pg
LOFAR is a SKA pathfinder instrument, which means that it serves not only as a cutting-edge instrument in its own right, but does so with the explicit aim of serving as testing grounds for technologies \& techniques which could be usefully implemented in the SKA. It is in this context, for example, that trailblazers such as NenuFAR \citepads{2012sf2a.conf..687Z}, a low-frequency extension of LOFAR, are tested. LOFAR is an interferometric array, with each of its elements made up of antennas combined into a phased array, themselves forming an interferometer. It thus consists of antennas which are combined to form stations, which are themselves distributed throughout the Netherlands and Europe.
\begin{figure}[h!] 
  \begin{minipage}[c]{0.45\linewidth}
    \centering
    \includegraphics[width=\linewidth]{images/superterp.png}
	\subcaption{\label{fig.lofar.superterp} LOFAR core, known as the Superterp.}
%    \vspace{4ex}
  \end{minipage}
  \hfill
  \begin{minipage}[c]{0.45\linewidth}
    \centering
    \includegraphics[width=.5\linewidth]{images/{lofar-core-map_grey.jpg}}
    \subcaption{\label{fig.lofar.core} Distribution of the so-called LOFAR "core" stations, which include the Superterp.}
%    \vspace{4ex}
  \end{minipage} 
  \begin{minipage}[c]{0.45\linewidth}
    \centering
    \includegraphics[width=\linewidth]{images/{Distribution_Remote_Stations}} 
    \subcaption{\label{fig.lofar.remote} LOFAR with both "core" and "remote" stations.}
%    \vspace{4ex}
  \end{minipage}
  \begin{minipage}[c]{0.45\linewidth}
    \centering
    \includegraphics[width=.9\linewidth]{images/{Distribution_International_Stations}} 
    \subcaption{\label{fig.lofar.international} International LOFAR. Newer stations (one in Ireland, three in Poland) are not shown here.}
%    \vspace{4ex}
  \end{minipage} 
\caption{\label{fig.lofar.distribution} Geographic location and distribution of LOFAR stations, explicitly showing what is meant by Superterp, core, remote and international stations. All images from \href{https://www.astron.nl/radio-observatory/astronomers/users/technical-information/lofar-array-configuration/lofar-array-conf}{the official ASTRON website}: \url{https://www.astron.nl/radio-observatory/astronomers/users/technical-information/lofar-array-configuration/lofar-array-conf}}
\end{figure}

@@

\pg
We see the presence, in both cases, of two very different antenna types. One of these antenna types is not a single antenna, but rather a phased array: 16 antenna dipoles distributed in a $4 \times 4$ array. International LOFAR stations include 96 such tiles, distributed in a ``filled-disk" pattern, where the disk fill ratio is determined by the size of the tile. Dutch ``remote" stations have 48 such tiles spread into 2 disks (HBA\textunderscore INNER and HBA\textunderscore OUTER), while ``core" station HBA tiles are split into two 24-tile crosses. This pattern is shown in \cref{fig.hba.tile}. The antennas from each tile are combined into a phased array with a single ``tile beam", and all tile beams are themselves combined as a phased array into a station beam\footnote{Reference: \href{https://www.astron.nl/radio-observatory/astronomers/technical-information/antennae/antennae-description}{ASTRON technical description}: \url{https://www.astron.nl/radio-observatory/astronomers/technical-information/antennae/antennae-description}}.% This station beam is then pointed digitally.
\begin{figure}[h!]
\includegraphics[width=0.45\textwidth]{images/{Schematic-diagram-of-a-24-tile-LOFAR-HBA-station-A-tile-is-made-of-16-dual-polarization.png}}
\caption{\label{fig.hba.tile} Layout of a single HBA tile. Source: \href{https://www.researchgate.net/profile/Sarod_Yatawatta/publication/281316394/figure/fig1/AS:614002589720576@1523401033149/Schematic-diagram-of-a-24-tile-LOFAR-HBA-station-A-tile-is-made-of-16-dual-polarization.png}{Sarod Yatawatta researchgate profile}}
\end{figure}

\pg
LBA antennas, use a very simple - and cost-effective - design. A typical LBA antenna is shown in \cref{fig.lba.ant}. 96 such antennas are spread in a semi-random pattern in each LOFAR station\footnote{Reference: \href{http://www.lofar.org/about-lofar/system/lofar-numbers/lofar-numbers}{LOFAR.org website}: \url{http://www.lofar.org/about-lofar/system/lofar-numbers/lofar-numbers}}. In Dutch stations, 48 of these are used at all times. They are combined as a single phased array.
\begin{figure}[h!]
\includegraphics[width=0.45\textwidth]{images/{lba.png}}
\caption{\label{fig.lba.ant} Picture of a single LBA antenna. Source: \href{https://i2.wp.com/lofar.ie/wp-content/uploads/2017/04/lba.png?resize=1500\%2C1000}{LOFAR technology website.}}
\end{figure}
\pg
LOFAR-HBA is sensitive to higher frequencies, from 120 MHz to 240 MHz\footnote{Reference: \href{http://www.lofar.org/about-lofar/system/lofar-numbers/lofar-numbers}{LOFAR.org website}: \url{http://www.lofar.org/about-lofar/system/lofar-numbers/lofar-numbers}}. It has a smaller field of view than LOFAR-LBA, but a better angular resolution. 
\pg
The dipole design frees observers from the need to physically point antennas at all: the final station pointing is achieved by digitally introducing delays in the recorded phase before averaging the data. In this sense, the pointing is achieved in a similar way as individual HBA tile pointing, albeit . The LBA antennas are receptive to signals emitted in the 30-80 MHz frequency range.
